Gringo 2.0

### 2008-07-28 doc/documentation/paper.tex
added some of svens remarks to the documentation
git-svn-id: https://potassco.svn.sourceforge.net/svnroot/potassco/branches/gringo2@411 3e02f596-3131-0410-8352-e04eeaec37cc

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
Gringo2.0 is a state of the art and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
The goal of this work is to give an overview of the features supported by gringo2.0. First, the general restrictions to the input language are stated. A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder support such a rich featured, less restricted input language.
There exist several output formats which can be used be different grounders and even a human readable format for debugging purposes.
Gringo also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
Furthermore gringo can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.

@@

\section{Interface}
Gringo can be started on the command line and has some parameters that can be useful.
The first parameters are gringo parameters and and arbritary number of filenames of the logic program.
\paragraph{--help}
Gives a short overview of all parameters.
\paragraph{--const}
You can set the value of constants defined in you logic program via the command line of gringo.
After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant. The integer value can even be an arithmetic expression. For example:
\begin{verbatim}
 gringo --const n=5*2
\end{verbatim}
the atom $n$ is replaced by $10$ in the logic program.

\paragraph{--convert}
If this parameter is used, the grounder expects an already grounded program as input.
For converting an already grounded logic program into an output format supported by gringo this is the right option. This mode saves a lot of memory but does not ground the program.
\paragraph{-l}
This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
This is the current default output format of gringo.
Sum aggregates with negative weights are handled like weight constraint in lparse for this output format. So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{-c}
In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answersets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line. Therefore adouble minus ``--'' must be given after all gringo related parameters are set. After them die clasp parameters follow.
For example:
\begin{verbatim}
 gringo -c --const n=5 -- --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
\paragraph{-p}
Parameter ``p'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
\paragraph{-g X}
This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. The X can be any number from $1$ to $7$ to refer to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7). If no normalform is specified, conglomeration(7)  is assumed.

@@

An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
It has to start with a lower case letter or an underscore.
An atom can have several parameters, encapsulated by parenthesis and separated with commas.
The parameters are either constants, variables or function symbols.
A function symbols do have the same syntax as atoms. This way they can be finitely nested.
As an example, an atom with three parameters is given:
The first parameter being a constant, the others being function symbols.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}

\paragraph{Rule}
A rule normally consists of a head and a body and ands with a dot.
The body is separated from the using ``\:\-'' as in
\begin{verbatim}
 head :- body.
\end{verbatim}
The body is a comma separated conjunction of expressions.
The supported expressions are introduced in the next section.
The variables used in the head must be restricted to a domain in the body.
Check (\ref{domainrestricted}) to see the restrictions on domains.
\newline
A rule can occur without a body as a fact, therefore the body is omitted.
\begin{verbatim}
 head.
\end{verbatim}
The same can be done with the body to express an integrity constraint.
\begin{verbatim}
 :- body.
\end{verbatim}

@@

\begin{itemize}
 \item 
\begin{verbatim}
 "Warning: sum with negative bounds in the body of a rule"
\end{verbatim}
This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp.
 \item 
\begin{verbatim}
 "Warning: multiple definitions of #const n"
\end{verbatim}
This warning is printed if more than one \#const statement is used to set the same constant.
The last attached value is used for grounding.
\item
\begin{verbatim}
 "Warning: p/3 is never defined"
\end{verbatim}
This warnings states that an atom is used in the body of a rule that never occurs in the head.
``p'' is the name of the atom and ``3'' is the arity.
\item
\begin{verbatim}
 "got ground program i hope u have enough memory :)"
\end{verbatim}
Use the command line option ``--convert'' to convert a ground program into another format.

@@

\paragraph{Errors}
Gringo2.0 stops the execution and prints an error mesaage to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of atoms that are not omega restricted.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted varibles: { ... }"
\end{verbatim}
This error message gives a list of atoms that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format.
\begin{verbatim}
 "Disjunction not allowed on this normalform."
 "Count aggregate/cardinality constraint not allowed
  in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with non trivial upper bound in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with non trivial upper bound in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Max aggregate not allowed in this normalform."
 "Max aggregate not allowed
  with negative weights in this normalform."
 "Min aggregate not allowed in this normalform."
 "Min aggregate not allowed 
  with negative weights in this normalform."
 "Optimize statement not allowed in this normalform."
\end{verbatim}
%\begin{verbatim}
% "Times aggregate not allowed in this normalform."
% "Times aggregate not allowed
%  with negative weights in this normalform."
%\end{verbatim}
These errors belong to the ASPils output format. Another normal form as to be chosen.
\end{itemize}

\subsection{Meta Constructs}
The ``\#'' sign in front of all meta expressions is optional and can therefore be omitted.
\paragraph{hide}
The hide construct is used to hide atoms from the output of the ASP solver.
Two different syntaxes are supported by gringo.
The first 
\begin{verbatim}
 #hide name/arity.
\end{verbatim}
and the second
\begin{verbatim}
 #hide name(X1, ... , Xn).
\end{verbatim}
Both variants hide the predicate ``name''.
The arity of the predicate can be given explicit or implicit with a list of variables.
The variables are just placeholder and do not unify. The second syntax is implemented for compatibility reasons only.
If the predicate that has to be hidden has arity zero, meaning it is an atom, the syntax is
\begin{verbatim}
 #hide name.
\end{verbatim}
Not to give explicitly all predicates that are hidden, only the ones which are visible the statement
\begin{verbatim}
 #hide.
\end{verbatim}
in conjunction with the ``show'' statement is used.
The ``show'' statement has the same syntax as ``hide'' but uses the keyword ``show''.
\begin{verbatim}
 #hide.
 #show p/3.
\end{verbatim}
The example hides all atoms except the predicate $p$ with arity $3$.
\paragraph{const}
Within the ``const'' statement, a default value for constants can be defined.
Constants are atoms used in the logic program and are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``T'' is always bound to the domain of ``time''.
This has the same effect as adding the domain ``time(T)'' to every rule the variable occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
Is has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answerset solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negative atoms do not have to be in the models of the program.
The compute statement is solver dependent and supported by every solver.

\subsection{How to write a logic program}
Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
Therefore the problem must be encoded as a set of rules.
Typically these are divided into a problem description and an instance of the problem.
The solution to a problem can then be read from answer sets of the rules.
The procedure of writing a logic program shall now be explained with a little example.
We want to write a problem encoding for the 3-colorability problem of graph coloring.
Given an arbitraty graph, no two connected edges may have the same color.
\newline
Usually you start with the non instance domains, in our example the colors we use.
\begin{verbatim}
 color(red). color(green). color(blue).
\end{verbatim}
After that, we have to color the vertices, but we must give attention that each vertex has only one color.
We can do this by simply saying that a vertex is colored with one color X, and that the same vertex does not have another color Y.
\begin{verbatim}
 colored(V,C) :- not colored(V,D), C != D, color(C;D), vertex(V).
\end{verbatim}
As you can see, you have to give the domain of the variables in the rule.
The variable V is used for vertices and the variables C and D are used for colors.
After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
We now need to restrict our solutions again by saying that two vertices that are connected by an edge should not have the same color.
We can do this by using an integrity constraint.
An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
\begin{verbatim}
 :- edge(V,U), colored(V,C), colored(U,C), vertex(V;U), color(C).
\end{verbatim}
Here we have the edge between two vertices V and U, and if both are colored with the same color C the assignment is not a solution to our problem.
\newline
We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
We just need to define several vertices and edges that do connect the vertices.
\begin{verbatim}
 vertices(1..4).
 edge(1,2). edge(1,4). edge(2,4).
 edge(3,1). edge(4,1). edge(3,2).
\end{verbatim}

If we give the problem description and instance to the grounder, it will generate a variable free version of the problem. This can then be solved by a solver and in the answer sets of the solver the ``colored'' predicate states a valid coloring of all vertices.
\newline
To ease the process writing logic programs some features in the input syntax were introduced and are now explained in detail.
\subsection{Features}
The extended features of the input syntax are now explained in detail. For a full syntax description in EBNF see appendix (\ref{appendix}).
\paragraph{Expanding Terms}
\subparagraph{Range Operator}
To define several atoms that stretch on one or more parameters over a certain range, two dots can used to express an increasing integer range.
Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
\begin{verbatim}
 times(1..3, 1..2).
\end{verbatim}
will result in the facts
\begin{verbatim}
 times(1,1). times(2,1). times(3,1).
 times(1,2). times(2,2). times(3,2).
\end{verbatim}

Using the range operator in the body of a rule, a disjunction of all members in the range is created. So the rules will be duplicated for every member.
\begin{verbatim}
 a :- p(1..3).
\end{verbatim}
will result in the rules
\begin{verbatim}
 a :- p(1).
 a :- p(2).
 a :- p(3).
\end{verbatim}
If the range operator is used in an aggregate, it creates a a comma separated list of atoms, so
\begin{verbatim}
 a :- {b(1..3)}.
\end{verbatim}
results in
\begin{verbatim}
 a :- {b(1), b(2), b(3)}.
\end{verbatim}

@@

\paragraph{Arithmetic Expressions}
The build in arithmetics can be used for integer arithmetic.
As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported. The bitwise operators are executed on signed integers using the binary complement for negative numbers.
Arithmetic expressions can occur everywhere in head and in the body, even in complex constructs.
As parameters, integer constants or expressions that evaluate to integers are used.
The simple example of using operators and range terms
\begin{verbatim}
 a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
 b(1,2).
 c(4).
\end{verbatim}
would result in the grounded program\footnote{gringo would simplify the grounded rules to facts, since all literals in the body are facts}
\begin{verbatim}
 a(0) :- b(1,2), c(4).
 a(1) :- b(1,2), c(4).
 a(2) :- b(1,2), c(4).
 b(1,2).
 c(4).
\end{verbatim}

\subparagraph{Relational Operators}
The relational operators are ``less than'' ($<$),\\ ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
They can be used to constrain certain variables.
For example, some rule is only generated for all unifications where X is greater than Y-1.
\begin{verbatim}
 a(X) :- b(X,Y), X >= Y-1.
 b(0,2).
 b(4,2).
\end{verbatim}
This would let $b(4,2)$ a valid substitution, since $4 >= 2-1$ but $b(0,2)$ not, so the grounded program could be:
\begin{verbatim}
 a(4) :- b(4,2).
 b(0,2).
 b(4,2).
\end{verbatim}

@@

\paragraph{True Negation}
True negation can be used. Therefore the atom is preceded by a minus sign.
\begin{verbatim}
 -a :- not  a.
  a :- not -a. 
\end{verbatim}
\paragraph{Disjunction}
A disjunction can be expressed using the pipe symbol ``$\mid$''.
A disjunction can only occur in the head of a rule.
\begin{verbatim}
 red | green | blue :- colored.
\end{verbatim}
Using disjunctions in logic programs increase expressiveness and complexity of the logic program.
A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.

>>>>

\begin{document}
\maketitle
\tableofcontents
\section{Introduction}
Answer Set Programming (ASP) \cite{baral02a} is a declarative approach for modeling and solving search problems, represented as logic programs.
An ASP system usually consists of a grounder and a solver.
First, a grounder translates a non-ground problem description into a propositional program, which can be processed by the second component of the system, the solver. This is illustrated in figure (\ref{fig:ASP}).
\begin{figure}[h]
\[
\xymatrix{*+[F]\txt{Problem\\representation}\ar[rr]^{Grounder}&&
		*+[F]\txt{Intermediate\\ representation} \ar[rr]^{Solver}&&
		*+[F]\txt{Variable\\ assignment}}
\]
\caption{Architecture of an ASP System}
\label{fig:ASP}
\end{figure}
gringo2.0 is a state of the art grounder and has many features that makes writing logic programs easy.
It was developed by Roland Kaminski and is orientated on the ideas of gringo1.0.
The goal of this work is to give an overview of the features supported by gringo2.0. First, the general restrictions to the input language are stated. A broad set of potentialities for writing encodings for logic programs exists. Currently, no other grounder supports such a rich featured, less restricted input language.
gringo2.0 supports several output formats and even a human readable format for debugging purposes.
It also has an internal interface to the solver clasp \cite{gekanesc07b}, so the output can directly by parsed to the solver and answersets can be computed.
Furthermore gringo can be used with clasp incrementally. Detailed information about incremental solving can be found here \cite{IncrementalASP} 
The syntax of the gringo2.0 input language can be found in the appendix \ref{appendix}.

@@

\section{Interface}
Gringo can be started on the command line and has some parameters that can be useful.
The first parameters are gringo specific parameters and any arbritary number of filenames of the logic program.
\paragraph{--help}
Gives a short overview of all parameters.
\paragraph{--const}
You can set the value of constants defined in your logic program via the command line of gringo.
After the parameter ``--const'' you have to give the name of the constant you want to set followed by a ``='' and the value of that constant. The integer value can even be an arithmetic expression. For example:
\begin{verbatim}
 gringo --const n=5*2
\end{verbatim}
the atom $n$ is replaced by $10$ in the logic program.

\paragraph{--convert}
If this parameter is used, the grounder expects an already grounded program as input.
For converting an already grounded logic program into an output format supported by gringo this is the right option. This mode saves a lot of memory but does not ground the program.
\paragraph{-l}
This mode prints smodels conform output. For a definition of the output format see \cite{lparseManual}.
Since some features of gringo are not natively supported for this output language, new atoms are introduced to create an equivalent program.\newline
This is the current default output format of gringo.
Sum aggregates with negative weights are handled like weight constraint in lparse for this output format. So a transformation as explaned in \cite{lparseManual} is used to get rid of the negative weights.
\paragraph{-c}
In this mode, the output is directly parsed to clasp, an ASP solver. For solving logic programs, no intermediate files are produced and the answersets are printed directly.
With this option enabled, all parameters for clasp are valid and can be declared in the command line. Therefore adouble minus ``--'' must be given after all gringo related parameters are set. After them die clasp parameters follow.
For example:
\begin{verbatim}
 gringo -c --const n=5 -- --supp-models 3
\end{verbatim}
Furthermore all restriction on smodels output hold as clasps internal interface currently reads smodels format.
\paragraph{-p}
Parameter ``p'' invokes the plain output method, so the output is printed in human readable format, which is also the input format for gringo.
\paragraph{-g [1-7]}
This parameter invokes the new output format for the intermediate languages ASPils \cite{atleastthisreferenceishouldgive}. The optional number can range from $1$ to $7$ and refers to one of the following normalforms (Simple=1, SModels=2, DisjunctiveLogicProgram=3, CModels=4, CModelsExtended=5, DLV=6, Conglomeration=7). If no normalform is specified, conglomeration(7)  is assumed.

@@

An atom is a predicate with a name. The name can consist of letters, digits and the underscore.
It has to start with a lower case letter or an underscore.
An atom can have several parameters, encapsulated by parenthesis and separated with commas.
The parameters are either constants, variables or function symbols.
Function symbols do have the same syntax as atoms. This way they can be finitely nested.
As an example, an atom with three parameters is given:
The first parameter being a constant $a$, the others being function symbols $f(b,X)$ and $g(f(X), Y))$.
\begin{verbatim}
 p(a,f(b,X), g(f(X),Y)).
\end{verbatim}

\paragraph{Rule}
A rule normally consists of a head and a body and ends with a dot.
The body is separated from the head using ``:-'' as in
\begin{verbatim}
 head :- body.
\end{verbatim}
The body is a comma separated conjunction of expressions.
The supported expressions are introduced in the next section.
The variables used in the head must be restricted to a domain in the body.
Check (\ref{domainrestricted}) to see the restrictions on domains.
\newline
A rule can occur without a body as a fact, therefore the body is omitted.
\begin{verbatim}
 head.
\end{verbatim}
The same can be done with the body to express an integrity constraint.
\begin{verbatim}
 :- body.
\end{verbatim}

@@

\begin{itemize}
 \item 
\begin{verbatim}
 "Warning: sum with negative bounds in the body of a rule"
\end{verbatim}
This warning occurs if a sum aggregate with negative weights has been found and the output format smodels is used or the output is redirected to clasp. In this format negative bounds are not allowed.
 \item 
\begin{verbatim}
 "Warning: multiple definitions of #const n"
\end{verbatim}
This warning is printed if more than one \#const statement is used to set the same constant.
The last attached value is used for grounding.
\item
\begin{verbatim}
 "Warning: p/3 is never defined"
\end{verbatim}
This warnings states that an atom is used in the body of a rule that never occurs in the head.
``p'' is the name of the predicate and ``3'' is the arity.
\item
\begin{verbatim}
 "got ground program i hope u have enough memory :)"
\end{verbatim}
Use the command line option ``--convert'' to convert a ground program into another format.

@@

\paragraph{Errors}
gringo2.0 stops the execution and prints an error message to standard error output.
\begin{itemize}
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  non domain predicates : { ... }"
\end{verbatim}
This error message gives a list of predicates that are not domain restricted.
 \item 
\begin{verbatim}
 "the following rule cannot not be grounded,
  weakly restricted variables: { ... }"
\end{verbatim}
This error message gives a list of predicates that are not lambda restricted. 
\item
\begin{verbatim}
 "syntax error on line Y column Z unexpected token: X"
\end{verbatim}
The parser can not read the input format.
\begin{verbatim}
 "Disjunction not allowed on this normalform."
 "Count aggregate/cardinality constraint not allowed
  in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with non trivial upper bound in this normalform."
 "Count aggregate/cardinality constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with non trivial upper bound in this normalform."
 "Sum aggregate/weight constraint not allowed 
  with negative lower bound in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Sum aggregate/weight constraint not allowed
  with negative weights in this normalform."
 "Max aggregate not allowed in this normalform."
 "Max aggregate not allowed
  with negative weights in this normalform."
 "Min aggregate not allowed in this normalform."
 "Min aggregate not allowed 
  with negative weights in this normalform."
 "Optimize statement not allowed in this normalform."
\end{verbatim}
%\begin{verbatim}
% "Times aggregate not allowed in this normalform."
% "Times aggregate not allowed
%  with negative weights in this normalform."
%\end{verbatim}
These errors belong to the ASPils output format. Another normal form as to be chosen.
\end{itemize}

\subsection{Meta Constructs}
The ``\#'' sign in front of all meta expressions is optional and can therefore be omitted.
\paragraph{hide}
The hide construct is used to hide atoms from the output of the ASP solver.
Two different syntaxes are supported by gringo.
The first 
\begin{verbatim}
 #hide name/arity.
\end{verbatim}
and the second
\begin{verbatim}
 #hide name(X1, ... , Xn).
\end{verbatim}
Both variants hide the predicate ``name''.
The arity of the predicate can be given explicit or implicit with a list of variables.
The variables are just placeholder and do not unify with other variables. For example hide p(X,X) would also hide the fact p(a,b). The second syntax is implemented for compatibility reasons only.
If the predicate that has to be hidden has arity zero the syntax is
\begin{verbatim}
 #hide name.
\end{verbatim}
Not to give explicitly all predicates that are hidden, only the ones which are visible the statement
\begin{verbatim}
 #hide.
\end{verbatim}
in conjunction with the ``show'' statement is used.
The ``show'' statement has the same syntax as ``hide'' but uses the keyword ``show''.
\begin{verbatim}
 #hide.
 #show p/3.
\end{verbatim}
The example hides all atoms except the predicate $p$ with arity $3$.
\paragraph{const}
Within the ``const'' statement, a default value for constants can be defined in the logic program.
while they are normally defined via the command line options.
A default value for a constant $n$ can be easily be set.
\begin{verbatim}
 #const n=5.
\end{verbatim}
\paragraph{domain}
A domain statement states that a specific variable has a global domain restriction.
The statement
\begin{verbatim}
 #domain time(T).
\end{verbatim}
makes sure that the variable ``T'' is always bound to the domain of ``time''.
This has the same effect as adding the domain ``time(T)'' to every rule the variable ``T'' occurs in.
\paragraph{compute}
The compute statement is a hint for the solver which models to compute.
Is has the syntax
\begin{verbatim}
 #compute number {a1, ..., aN, not aN+1, ..., not aM}.
\end{verbatim}
$number$ is the number of models that shall be computed by the answerset solver and can be omitted. If it is zero, no restriction to the number of models is applied.
All positive atoms in the compute statement must be in the computed models, and all negative atoms do not have to be in the models of the program.
The compute statement is solver dependent and supported by every solver.

\subsection{How to write a logic program}
Writing a logic program to solve a problem is rather describing a problem than describing an algorithm to solve a problem.
Therefore the problem must be encoded as a set of rules.
Typically these are divided into a problem description and an instance of the problem.
The solution to a problem can then be read from the answer sets of the program.
The procedure of writing a logic program shall now be explained with a little example.
We want to write a problem encoding for the 3-colorability problem of graph coloring.
Given an arbitraty graph, no two connected edges may have the same color.
\newline
Usually you start with the non instance domains, in our example the colors we use.
\begin{verbatim}
 color(red). color(green). color(blue).
\end{verbatim}
After that, we have to color the vertices, but we must give attention that each vertex has only one color.
We can do this by simply saying that a vertex is colored with one color X, and that the same vertex does not have another color Y.
\begin{verbatim}
 colored(V,C) :- not colored(V,D), C != D, color(C;D), vertex(V).
\end{verbatim}
As you can see, you have to give the domain of the variables in the rule.
The variable V is used for vertices and the variables C and D are used for colors.
After every vertex now has a color we would get all possible solutions of all vertices colorfully painted.
We now need to restrict our solutions again by saying that two vertices that are connected by an edge should not have the same color.
We can do this by using an integrity constraint.
An integrity constraint is a rule with an empty head, and if an assignment of variables fulfills the body, the assignment is not a solution of our problem.
\begin{verbatim}
 :- edge(V,U), colored(V,C), colored(U,C), vertex(V;U), color(C).
\end{verbatim}
Here we have the edge between two vertices V and U, and if both are colored with the same color C the assignment is not a solution to our problem.
\newline
We have finally written our problem description, all we need now is a problem instance. This step is fairly simple.
We just need to define several vertices and edges that do connect the vertices.
\begin{verbatim}
 vertices(1..4).
 edge(1,2). edge(1,4). edge(2,4).
 edge(3,1). edge(4,1). edge(3,2).
\end{verbatim}

If we give the problem description and instance to the grounder, it will generate a variable free version of the problem. This can then be solved by a solver and in the answer sets of the solver the ``colored'' predicate states a valid coloring of all vertices.
\newline
To ease the process writing logic programs some features in the input syntax were introduced and are now explained in detail.
\subsection{Features}
The extended features of the input syntax are now explained in detail. For a full syntax description in EBNF see appendix (\ref{appendix}).
\paragraph{Expanding Terms}
\subparagraph{Range Operator}
To define several atoms that stretch on one or more parameters over a certain range, two dots can be used to express an increasing integer range.
Using the range operator in the head of a rule, it works as a conjunction of all the members in the range, so for every number the rule is duplicated.
\begin{verbatim}
 times(1..3, 1..2).
\end{verbatim}
will result in the facts
\begin{verbatim}
 times(1,1). times(2,1). times(3,1).
 times(1,2). times(2,2). times(3,2).
\end{verbatim}

Using the range operator in the body of a rule, a disjunction of all members in the range is created. So the rules will be duplicated for every member.
\begin{verbatim}
 a :- p(1..3).
\end{verbatim}
will result in the rules\footnote{gringo would simplify the grounded rules to facts, since all literals in the body are facts}
\begin{verbatim}
 a :- p(1).
 a :- p(2).
 a :- p(3).
\end{verbatim}
If the range operator is used in an aggregate, it creates a a comma separated list of atoms, so
\begin{verbatim}
 a :- {b(1..3)}.
\end{verbatim}
results in
\begin{verbatim}
 a :- {b(1), b(2), b(3)}.
\end{verbatim}

@@

\paragraph{Arithmetic Expressions}
The build-in arithmetics can be used for integer arithmetic.
As operators addition (+), subtraction (-), multiplication (*), division (/), modulo (mod) and the absolute value operator (abs) can be used.
Furthermore the bitwise operators and(\&), or(?), xor(\textasciicircum) and invert($\sim$) are supported. The bitwise operators are executed on signed integers using the binary complement for negative numbers.
Arithmetic expressions can occur everywhere in head and in the body, even in complex constructs.
As parameters, integer constants or expressions that evaluate to integers are used.
The simple example of using operators and range terms
\begin{verbatim}
 a(X-1 .. X+1) :- b(X,Y), c(abs(Y * 2 * -X)).
 b(1,2).
 c(4).
\end{verbatim}
would result in the grounded program
\begin{verbatim}
 a(0) :- b(1,2), c(4).
 a(1) :- b(1,2), c(4).
 a(2) :- b(1,2), c(4).
 b(1,2).
 c(4).
\end{verbatim}

\subparagraph{Relational Operators}
The relational operators are ``less than'' ($<$),\\ ``greater than'' ($>$), ``less equal'' ($<=$), ``greater equal'' ($>=$) , ``equal'' ($==$) and ``not equal'' ($!=$).
They can be used to constrain certain variables.
For example, the following rule is only generated for all unifications where X is greater than Y-1.
\begin{verbatim}
 a(X) :- b(X,Y), X >= Y-1.
 b(0,2).
 b(4,2).
\end{verbatim}
This would let $b(4,2)$ a valid substitution, since $4 >= 2-1$ but $b(0,2)$ not, so the grounded program could be:
\begin{verbatim}
 a(4) :- b(4,2).
 b(0,2).
 b(4,2).
\end{verbatim}

@@

\paragraph{Classical Negation}
True negation can be used. Therefore the atom is preceded by a minus sign.
\begin{verbatim}
 -a :- not  a.
  a :- not -a. 
\end{verbatim}
\paragraph{Disjunction}
A disjunction can be expressed using the pipe symbol ``$\mid$''.
A disjunction can only occur in the head of a rule.
\begin{verbatim}
 red | green | blue :- colored.
\end{verbatim}
Using disjunctions in logic programs increase expressiveness and complexity of the logic program.
A solver for disjunctive logic programs such as claspD \cite{claspD} or cmodels \cite{cmodels} is needed to solve such programs.
